Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-08-16 manuscript.qmd
use full graph for fig. 2
Related: #11

The transition matrix $\mathbf{A}$ is parametrized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network. Directed arcs from the primary tumor to a LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Directed arcs from a LNL $v$ to a LNL $r$ are associated with a probability $t_{vr}$ for the tumor to progress from one LNL to the next. In this paper, we build on the DAG shown in @fig-small-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].

![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region. Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables. Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step. The orange arcs represent the sensitivity and specificity of the observational modality (e.g. CT, MRI, pathology, ...).](static/small-graph.svg){#fig-small-graph width=40%}

@@

The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-small-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors. In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories. Therefore, the parameter space of the unilateral model is:

@@

1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-small-graph}.
2. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha \in [0,1]$ that captures a "degree of asymmetry": \
   $$
   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
   $$ {#eq-mixing}
   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline. Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.

>>>>

The transition matrix $\mathbf{A}$ is parametrized using a directed acyclic graph (DAG) as an abstract representation of the underlying lymphatic network. Directed arcs from the primary tumor to a LNL are associated with a probability $b_v$ for direct spread to LNL $v$ during one time step. Directed arcs from a LNL $v$ to a LNL $r$ are associated with a probability $t_{vr}$ for the tumor to progress from one LNL to the next. In this paper, we build on the DAG shown in @fig-full-graph which was obtained by maximizing the model evidence as described in [cite 2024 graph extension paper].

![Directed acyclic graph (DAG) representing the abstract lymphatic network in the head and neck region. Blue nodes are the LNLs' hidden random variables, the red node represents the tumor, and the orange square nodes depict the binary observed variables. Red and blue arcs symbolize the probability of lymphatic spread along that edge during one time-step. The orange arcs represent the sensitivity and specificity of the observational modality (e.g. CT, MRI, pathology, ...).](static/full-graph.svg){#fig-full-graph width=40%}

@@

The single-patient likelihood $\ell$ in @eq-single-patient-llh depends on the spread parameters shown in @fig-full-graph via the transition matrix $\mathbf{A}$ and on the binomial parameters $p_{\text{T}x}$ via time-priors. In this work, we will only differentiate between "early" (T1 & T2) and "advanced" (T3 & T4) T-categories. Therefore, the parameter space of the unilateral model is:

@@

1. We assume that both ipsilateral and contralateral spread is described through the same graph shown in {#fig-full-graph}.
2. We assume the spread *among* the LNLs to be same on both sides. It is reasonable to assume the lymphatic system is symmetric. Thus, the spread rates from one LNL to the other should be symmetric, too. Formally, this means \
   $$
   \begin{aligned}
   b_v^\text{c} &\neq b_v^\text{i} \\
   t_{rv}^\text{c} &= t_{rv}^\text{i}
   \end{aligned}
   $$ {#eq-symmetries}
   for all $v \leq V$ and $r \in \operatorname{pa}(v)$.
3. The tumor's spread to the contralateral side in case of an extension over the midline is larger than if it was clearly lateralized, but smaller than its spread to the ipsilateral side. This assumption stems from a simple thought experiment: Consider moving the tumor from a clearly lateralized position accross the mid-sagittal plane to the same position, but on the contralateral side. In the beginning we would have $b_v^\text{c} < b_v^\text{i}$, while in the end, the situation is reversed. If a tumor extends over the mid-sagittal line, its contralateral spread rate can be expected to be in between these two extremes. We encode this in a *mixing parameter* $\alpha \in [0,1]$ that captures a "degree of asymmetry": \
   $$
   b_v^{\text{c},\epsilon=\texttt{True}} = \alpha \cdot b_v^\text{i} + (1 - \alpha) \cdot b_v^{\text{c},\epsilon=\texttt{False}}
   $$ {#eq-mixing}
   This means the model now uses three different sets of parameters to describe the spread from the tumor to the LNLs: $b^\text{i}_v$ for the spread to the ipsilateral LNLs, $b_v^{\text{c},\epsilon=\texttt{False}}$ for the spread to the contralateral LNLs as long as the tumor is clearly lateralized, and finally $b_v^{\text{c},\epsilon=\texttt{True}}$ when it crosses the midline. Note, however, that these three sets of spread rates only account for $2 \cdot 2^V + 1$ parameters, since they are coupled via the mixing parameter $\alpha$.
